### 2023-02-04 introduction.tex
Additional comments and changes by Reed

\begin{figure}[h!]
% \begin{framed}
    \begin{minipage}[c]{0.65\textwidth}
        \includegraphics[scale=1]{fig-aln.pdf}
    \end{minipage}\hfill
    \begin{minipage}[c]{0.35\textwidth}
        \caption{
        \textbf{Standard algorithms produce suboptimal alignments.}
        (a) shows a possible alignment of an ancestor (A) sequence and a descendant (D) sequence.
        (b), (c), and (d) are the results of different aligners.
        Nucleotide mismatches are highlighted in red, phase zero, one, and two indels are shown in gray, purple, and orange, respectively.
        Phase one indel (purple) is also of type I (synonymous) while phase two indel (orange) is of type II (nonsynonymous). 
        COATi is able to retrieve the biological alignment (highlighted in blue).
        }
    \label{fig:aln}
    \end{minipage}
% \end{framed}
\end{figure}

>>>>

% Figures are 86 mm or 178 mm wide
\begin{figure}[h!]
    \centering%
    \includegraphics[scale=1]{fig-aln.pdf}
    \par
    \caption{
        Standard algorithms produce suboptimal alignments.
        (a) shows a possible alignment of an ancestor sequence (A) and a descendant sequence (D).
        (b), (c), and (d) are the results of different aligners.
        Nucleotide mismatches are highlighted in red. Phase 0, phase 1, and phase 2 indels are shown in gray, purple, and orange, respectively.
        Additionally, the orange indel is type II (an amino-acid indel plus an amino-acid change) while the purple indel is type I (an amino-acid indel only).
        COATi is the only aligner able to retrieve the biological alignment in this example.
        }
    \label{fig:aln}
\end{figure}

### 2023-02-04 materials_methods.tex
Additional comments and changes by Reed

Genomes for model organisms are often of high quality after being refined over many iterations and having their coding sequences meticulously curated.
On the contrary, non-model organisms typically have lower-quality genomes that have been only partially curated.
Low-quality genomes often lack the amount of sequencing data needed to fix artifacts, including missing exons, erroneous mutations, and indels \parencite{jackman2018tigmint}.
FSTs and their powerful methods provide a well-suited framework to statistically align a sequence from a non-model organism against a sequence from a model organism.

COATi implements the pairwise alignment of a low-quality sequence (descendant)  against a high-quality sequence (ancestor) as a path through the Evolution FST (Fig. \ref{fig:evolution-fst}), based on existing transducers (e.g. \cite{holmes2001evolutionary}).
This FST is the result of composing a substitution FST that encodes a codon model (Fig. \ref{fig:evolution-fst}-a) and an indel FST that models insertions and deletions, including frameshifts (Fig. \ref{fig:evolution-fst}-b).
A key innovation of this FST with respect to others is the combination of a codon substitution model with a nucleotide-based geometric indel model that allows gaps to occur at any position.

Composing both sequences with the Evolution FST results in the transducer of all possible alignments.
Any path through this FST represents a pairwise alignment, while the shortest path corresponds to the best alignment.
All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \parencite{allauzen2007openfst}.

\begin{figure}[h!]
\begin{framed}
\centering
    \includegraphics[width=\textwidth]{fig-evolution-fst.pdf}
    \caption{The Evolution FST is assembled by composing a substitution FST and
    an indel FST. Each node represents a state in an FST while arcs display
    possible transitions between states (and their weights). (a) The
    substitution FST encodes a 61x64 codon substitution model with $61\cdot 64$ arcs
    from M to S. (b) The indel FST allows for insertions (I to U) and deletions
    (D to V). Contiguous insertions and deletions are always arranged for
    insertions to precede deletions to limit equivalent alignments.}
    \label{fig:evolution-fst}
\end{framed}
\end{figure}

Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics.
COATi implements the Muse and Gaut (1994) codon model (codon-triplet-mg) and the Empirical Codon Model \parencite{kosiol_ECM_2007} (codon-triplet-ecm).
It also lets the user provide a codon substitution matrix.
The default FST model (codon-triplet-mg) does not allow substitutions from stop codons, although it supports mutations to (early) stop codons under the assumption that these are artifacts common in low-quality data.

COATi also features a marginal substitution model with probability matrix

\[P'_{ijp} = \mathlarger{\mathlarger{\sum}}_{cod} \thickspace \begin{cases}
P(i \thinspace | \thinspace cod) \enspace \text{if} \enspace cod_p = j\\
0 \enspace \text{otherwise}
\end{cases} \]

Where $P'_{ijp}$ represents the probability that codon $i$ from the ancestor
sequence changes to nucleotide $j$ of the descendant sequence at position $p$
$\in \{0, 1, 2\}$ of the reading frame.
$P$ is the substitution probability matrix Muse and Gaut or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg (default model) and codon-marginal-ecm.
These models emphasize the position where the substitution in a codon occurs, help restrict the effects of low-quality data in the descendant sequence, and allow more than one substitution per codon.
In combination with the indel model, alignment using the marginal model is implemented using dynamic programming.

>>>>

Genome quality impacts conclusions drawn from comparative genomic studies. \TODO{Juan, add citations.}
Genomes for model organisms often get refined over many iterations and achieve high quality with meticulously curated protein coding sequences.
In contrast, genomes for non-model organisms might only receive partial curation and typically have lower quality sequences and annotations.
These genomes often lack the amount of sequencing data needed to fix artifacts, including missing exons, erroneous mutations, and indels \parencite{jackman2018tigmint}.
FSTs and their powerful methods provide a well-suited framework to statistically align a sequence from a non-model organism against a sequence from a model organism.

COATi implements the pairwise alignment of a potentially lower-quality sequence against a high-quality sequence as a path through the Evolution FST (Fig. \ref{fig:evolution-fst}), based on existing transducers (e.g. \cite{holmes2001evolutionary}).
Here, COATi treats the high-quality (reference) sequence as the ``ancestor'' and the potentially lower-quality sequence as the ``descendant''.
This FST is the result of composing a substitution FST that encodes a codon model (Fig. \ref{fig:evolution-fst}-a) and an indel FST that models insertions and deletions, including frameshifts (Fig. \ref{fig:evolution-fst}-b).
A key innovation of this FST with respect to others is the combination of a codon substitution model with a nucleotide-based geometric indel model that allows gaps to occur at any position.

Composing both sequences with the Evolution FST results in the transducer of all possible alignments.
Any path through this FST represents a pairwise alignment, while the shortest path corresponds to the best alignment.
All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \parencite{allauzen2007openfst}.
However, the Evolution FST has a large state space to keep track of codon substitution rates when codons might be interspersed with indel events. This additional state space increase the computational complexity of the alignment algorithm.

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{fig-evolution-fst.pdf}
\par
\caption{The Evolution FST is assembled by composing a substitution FST and an indel FST.
Each node represents a state in an FST while arcs display possible transitions between states (and their weights). Unlabeled arcs have weights of 1.
(a) The substitution FST encodes a $61 \times 64 $ codon substitution model with 3904 arcs from M to S. These arcs consume three nucleotides from the input tape and emit three nucleotides to the output tape. The weight of each arc is a conditional probability derived from a codon substitution model. 
(b) The indel FST allows for insertions (I to U) and deletions (D to V). Insertion arcs are weighted according to the codon model's stationary distribution of nucleotides, and deletion arcs have a weight of 1. Contiguous insertions and deletions are always arranged for insertions to precede deletions to limit equivalent alignments.}
\label{fig:evolution-fst}
\end{figure}

Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics.
COATi implements the Muse and Gaut (1994) codon model (codon-triplet-mg) and the Empirical Codon Model \parencite{kosiol_ECM_2007} (codon-triplet-ecm).
It also lets the user provide a codon substitution matrix.
The default FST model (codon-triplet-mg) does not allow substitutions from stop codons \TODO{Juan is this still correct? Do we need to mention the extra error rate.}, although it supports mutations to (early) stop codons under the assumption that these are artifacts common in low-quality data.

To reduce the runtime complexity of COATi, we have also developed an approximation of the Evolution FST that can be implemented with standard dynamic programming techniques. This approximation uses a marginal substitution model where the output nucleotides are independent of one another and only depend on the input codon and position. This produces a $\left(61 \times 3 \right) \times 4$ substitution model and eliminates the need to track dependencies between output nucleotides.

A marginal substitution model is calculated from a standard substitution model by calculating the marginal probabilities that each ancestral codon produces specific descendant nucleotides at each reading frame positions.
%
Specifically, let
$P_\text{cod}\left(Y_0 \cdot Y_1 \cdot Y_2 \middle| X_0 \cdot X_1 \cdot X_2\right)$ represent transition probabilities from a standard codon model, and
%
\[
P_\text{mar}\left(Y_p = y \middle| X_0 \cdot X_1 \cdot X_2\right) =
\sum_{Y_0 \cdot Y_1 \cdot Y_2} I(Y_p = y) P_\text{cod}\left(Y_0 \cdot Y_1 \cdot Y_2 \middle| X_0 \cdot X_1 \cdot X_2\right)
\]
%
represent the marginal transition probabilities, where
$p \in \{0, 1, 2\}$ is the position of the descendent nucleotide relative to the ancestral reading frame.
COATi contains marginal models for both Muse and Gaut or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg (default model) and codon-marginal-ecm.
These models emphasize the position where the substitution in a codon occurs, help restrict the effects of low-quality data in the descendant sequence, and allow more than one substitution per codon.
In combination with the indel model, alignment using the marginal model is implemented using dynamic programming.

### 2023-02-04 results_discussion.tex
Additional comments and changes by Reed

\section*{Results and Discussion}
Using 16000 human genes and their gorilla homologous pairs from the ENSEMBL
database \parencite{ensembl_hubbard_2002} we simulated a data set of pairwise
alignments with empirical gap patterns.
We used the data set to evaluate the accuracy of popular cutting edge aligners
Clustal$\Omega$ v1.2.4 \parencite{clustal_omega_sievers_2011},
MACSE v2.06 \parencite{ranwez_macse_2011}, MAFFT v7.407
\parencite{mafft_katoh_2002}, and PRANK v.170427 \parencite{prank_loytynoja_2014}
together with COATi.

After downloading, sequences longer than 6000 nucleotides were filtered out
(2232), and 8369 alignments contained gaps identified by at least one
aligner.
We randomly introduced gap patterns extracted from all five methods into the
5399 initially ungapped sequence pairs to generate the benchmark alignments.
Alignment accuracy was measured using the distance metric $d_{seq}$
\parencite{metrics_blackburne_whelan_2011} between simulated and inferred
alignments.
In addition, accuracy of positive and negative selection was calculated
using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
\parencite{ka_ks_li_1993}.

% Software comparison table
\begin{table}[!ht]
\centering
 \frame{\input{figures/table-comp}}
 \caption{Accuracy of COATi, PRANK, MAFFT, Clustal$\Omega$, and MACSE, on 5399
         simulated sequence pairs. Perfect alignments have ($d_{seq}=0$), best
         alignments have lowest $d_{seq}$, and imperfect alignments have
         $d_{seq}>0$ when at least one aligner found a perfect alignment.}
 \label{table:comp}
\end{table}

COATi was significantly more accurate (lower $d_{seq}$) at inferring simulated alignments compared to other methods; all p-values were less than $2.2\cdot10-16$ according to the one-tailed Wilcoxon signed rank test.
In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately retrieved events of positive selection (Table \ref{table:comp}).
It obtained better results compared to a wide variety of alignment strategies.
Clustal$\Omega$, performing a common approach of aligning via amino acid translations, obtained the highest average alignment error and had difficulties retrieving positive selection.
MACSE, which allows frameshifts, is also based on an amino acid model and obtained similar results to the DNA-based MAFFT.
PRANK, using a codon model, had a similar average alignment error to MACSE and MAFFT but had issues recovering the simulated alignments.

>>>>

\section*{Results and Discussion}
Using 16000 human genes and their gorilla homologous pairs from the ENSEMBL
database \parencite{ensembl_hubbard_2002}, we simulated a data set of pairwise
alignments with empirical gap patterns.
We used the data set to evaluate the accuracy of popular cutting edge aligners
Clustal$\Omega$ v1.2.4 \parencite{clustal_omega_sievers_2011},
MACSE v2.06 \parencite{ranwez_macse_2011}, MAFFT v7.407
\parencite{mafft_katoh_2002}, and PRANK v.170427 \parencite{prank_loytynoja_2014}
together with COATi.

After downloading, we removed 2232 sequences longer than 6000 nucleotides, identified 8369 sequence pairs that contained gaps identified by at least one aligner, and 5399 ungapped sequence pairs.
We then randomly introduced gap patterns extracted from all five methods into the ungapped sequence pairs to generate the benchmark alignments.
Alignment accuracy was measured using the distance metric $d_{seq}$
\parencite{metrics_blackburne_whelan_2011} between simulated and inferred
alignments.
In addition, accuracy of positive and negative selection was calculated
using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
\parencite{ka_ks_li_1993}.

% Software comparison table
\begin{table}[!ht]
\centering
\input{figures/table-comp}
 \caption{COATi generates better alignments than other alignment algorithms. Results of COATi, PRANK, MAFFT, Clustal$\Omega$, and MACSE aligning 5399 empirically simulated sequence pairs. Perfect alignments have $d_{seq}=0$, best alignments have the lowest $d_{seq}$, and imperfect alignments have $d_{seq}>0$ when at least one aligner found a perfect alignment.}
 \label{table:comp}
\end{table}

COATi was significantly more accurate (lower $d_{seq}$) at inferring simulated alignments compared to other methods; all p-values were less than $2.2 \cdot 10^{-16}$ according to the one-tailed Wilcoxon signed rank test.
In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately retrieved events of positive selection (Table \ref{table:comp}).
It obtained better results compared to a wide variety of alignment strategies.
Clustal$\Omega$, performing a common approach of aligning via amino acid translations, obtained the highest average alignment error and had difficulties retrieving positive selection.
MACSE, which allows frameshifts, is also based on an amino acid model and obtained similar results to the DNA-based MAFFT.
PRANK, using a codon model, had a similar average alignment error to MACSE and MAFFT but had issues recovering the simulated alignments.
